---
title: "Read, Write and Cleaning Data"
author: "Pongsun B."
date: "2024-08-22"
output:
  html_document: default
  pdf_document: default
header-includes:
  - \pagenumbering{gobble}
---

## ==== First things first ====

Important matters should be dealt with before other things.

1.  Setting the working directory in R can be done using the `setwd()` function.

### Mac

`"/Users/USERNAME/Documents/myR/"`

### Windows

`"C:\Users\USERNAME\Documents\myR\"`

To read a CSV data file within a sub-folder `Data` under the working directory

`"./Data/FILENAME.csv"`

2.  Installing R packages can be done using the `install.packages()` function.

3.  Loading R packages is quite easy by just typing `library()` function.

4.  Installing and Loading R packages can be combined into one step by `if(!require()){install.packages(" ")}`

```{r Set working directory}
# Check current working directory  
getwd()
# Set working directory

# Call required packages, ex. "readxl", "tidyverse"
library(readxl)
library(tidyverse)
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(tidyverse)){install.packages("readxl")}
if(!require(tidyverse)){install.packages("dplyr")}
```

## ==== Read Data in R ====

### Read CSV (Comma Separated Value) file

```{r Read CSV file}
# Read CSV file "RenSteve.csv" with `read_csv`
ClassData <- read_csv("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/RenSteve.csv")
ClassData
# Use `spec()` to retrieve the full column specification for this data
spec(ClassData)
# Read CSV file with `read.csv` and `read.table`
ClassData2 <- read.csv("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/RenSteve.csv", header = F)
ClassData3 <-  read.table("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/RenSteve.csv", header = F, stringsAsFactors = T)
ClassData3
summary(ClassData3)
# Read CSV file from Dropbox (MUST edit dl=0 to dl=1!!)
# URL: https://www.dropbox.com/scl/fi/gg5okzhny0wjejrgw066s/sales.csv?rlkey=qggpmjmyfh2w8e0r3wt0tmsbd&dl=0
df_sales <- read_csv("https://www.dropbox.com/scl/fi/gg5okzhny0wjejrgw066s/sales.csv?rlkey=qggpmjmyfh2w8e0r3wt0tmsbd&dl=1")
df_sales
# Read shared file from OneDrive  (MUST edit "embed" to "download"!!)
# URL: https://onedrive.live.com/embed?resid=B83741508FA07690%211290&authkey=!AEBphFQulUSB8HQ
df_sales2 <- read_csv("https://onedrive.live.com/download?resid=B83741508FA07690%211290&authkey=!AEBphFQulUSB8HQ")
df_sales2
# Another method
if(!require(stringr)){install.packages("stringr")}
import_url <- str_replace("https://onedrive.live.com/embed?resid=B83741508FA07690%21259&authkey=!AKpzaWBcwCfeZ9E","embed","download")
test_Thai <- read_tsv(import_url)
test_Thai
# Another way is using "Import Dataset" feature of RStudio
# Import Dataset > From Text (base) or From Text (readr)

RenSteve2 <- read_csv("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/RenSteve.csv",
                      col_names = FALSE)
summary(RenSteve2)
```

### Read Excel file

```{r Read Excel file}
# Read Excel file "RenSteve.xlsx" with `read_excel`
ClassData4 <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/RenSteve.xlsx")
ClassData4
# Assign "range of data" to be read
# File: TopMovies.xlsx
# Range: C5:F55
TopMovies <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/TopMovies.xlsx", range = "C5:F55")
TopMovies
# Excel with multiple sheets
# File: multi.xlsx
df_drinks <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/multi.xlsx", sheet = "drinks")
df_drinks
df_accident <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/multi.xlsx", sheet = "accident")
df_accident
# Read Excel file from Dropbox
tempURL <- "https://www.dropbox.com/s/6un9tfrz6vy74fq/RenSteve.xlsx?dl=1"
download.file(tempURL, destfile = "local.xlsx")
my.data <- read_excel("local.xlsx", col_names = TRUE, col_types = "text")
multi <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/multi.xlsx")
```

## ==== Write Data in R ====

### Save data frame to CSV file

```{r Save data frame to CSV file}
# Save data frame to CSV file with "write.csv" or "write_csv"
# Use write.csv
write.csv(multi)
# Use write_csv
# Require library(readr)

```

### Write data frame to Excel file

<https://sparkbyexamples.com/r-programming/export-excel-in-r-xlsx-or-xls/><https://datatofish.com/export-dataframe-to-excel-in-r/>

The r-package `writexl` is required to write `.xlsx` files. More specifically, to write to an Excel file we will use the `write_xlsx` function.

Let's start by creating a data frame with some variables.

`df <- data.frame("Age" = c(21, 22, 20, 19, 18, 23), "Names" = c("Andreas", "George", "Steve", "Sarah", "Joanna", "Hanna"))`

`str(df)`

To write to `.xlsx` using the `write_xlsx` function from the `writexl` package:

`if(!require(writexl)){install.packages("writexl")}`

`write_xlsx(df, 'names_ages.xlsx')`

```{r Write data frame to Excel file}
df <- data.frame("Age" = c(21, 22, 20, 19, 18, 23), 
                 "Names" = c("Andreas", "George", "Steve", "Sarah", "Joanna", "Hanna"))
str(df)

# Export output to .xlsx file
if(!require(writexl)){install.packages("writexl")}
write_xlsx(df, 'names_ages.xlsx')

```

## ==== Data Cleaning in R ====

### How to delete a row of data frame?

To delete a row of data frame in R, you can use the `–` operator. For example, if you want to remove the first row from a data frame in R you can use the following code: `dataFrame <- dataFrame[-1, ]`. This code will remove the first row from the data frame. Of course, if you need to drop multiple rows, you can add them using the vector function: `dataFrame <- [-c(1, 2, 3),]`.

It is also possible to use the `slice()` function from the package `dplyr` to delete a row by its index.

```{r To delete a row of data frame}
# Read Excel file "play_data.xlsx" with "read_excel"
play_data <- read_excel("/Users/slick/Library/CloudStorage/OneDrive-ChulalongkornUniversity/University/Senior/Comp App Trans/Data/play_data.xlsx")
play_data
# Delete the second row
play_data <- play_data[-2, ]
# Delete the 5th to 7th rows 
play_data <- play_data[-c(5,6,7),]
play_data
# To remove 1st to 4th rows using the slice() function
play_data <- play_data[-c(1:4),]
play_data
```

### How to remove a row from a matrix?

Removing a row from a matrix can be done in the same way you delete a row from a data frame: `mtx <- mtx[-1,]`.

```{r To remove a row from a matrix}
mtx <- matrix(seq(1,15), byrow = TRUE, ncol = 3)
mtx
# Drop the last row from matrix
mtx <- mtx[-1,]
mtx
```

### Delete rows based on conditions using the `filter()` function

Sometimes we need to remove a row (or rows) based on one or many conditions. We can use the `filter()` function from the package `dplyr`.

Note that %\>% is called a "Pipeline", which means "then".

```{r using the filter() function}
play_data
# Delete the data with first name "James"
play_data2 <- play_data %>%
  filter(First_Name != "James")
play_data2
# Delete the data with first name "James" and all "fifth" day
play_data3 <- play_data %>%
  filter(First_Name != "James" & Day != "Fifth")

play_data3


```

### Remove ALL rows with missing values (NA)

To remove the missing values, use the `na.omit()` function.

`dataf_na <- na.omit(dataf_na)`

Alternatively, we can remove all rows with NA in the data frame using the `drop_na()` function in the package `tidyr`.

```{r Remove the missing values (NA)}
# Use the `na.omit()` function
play_data4 <- play_data %>%
  na.omit()
# Use the `drop_na()` function in the  package `tidyr`
play_data5 <- play_data %>%
  drop_na()

# Create test data
student.data <- data.frame(student = c("Ben","Jena","Jay","Lek","Bruce"),
                gender = c("M","F","F",NA,"M"),
                score1 = c(10,20,NA,40,50),
                score2 = c(25,NA,10,20,17),
                score3 = c(6,4,-99999,5,7))
student.data$score3[student.data$score3 == -99999] <- NA
student.data
# To remove all data with NA
new.student.data <- student.data %>%
  na.omit()
# Use the `complete.cases` function
student.data1 <- student.data[complete.cases(student.data), ]
  
student.data2 <- complete.cases(student.data)

# Check only "score1","score2" and "score3"
student.data3 <- student.data[complete.cases(student.data[, c("score1", "score2", "score3")]), ]


# To replace NA with the mean of known data
which(is.na(student.data$score1))
student.data$score1[which(is.na(student.data$score1))] <- mean(student.data$score1, na.rm = TRUE)


```

### How to remove duplicate rows and columns

To delete duplicate rows in R you can use the `duplicated()` function.

`study_df.un <- study_df[!duplicated(df), ]`

To remove duplicate columns we added the `as.list()` function and removed the `","`.

The function `distinct()` is used to keep only unique/distinct rows from the data frame. When working with the `distinct()` function, if there are duplicate rows, only the first row of the identical ones is preserved. It is also possible to delete duplicate rows based on values in a certain column.

```{r Remove duplicate rows and columns}
# Creating a data frame
example_df <- data.frame(FName = c('Steve', 'Steve', 'Erica', 'John', 'Brody', 'Lisa', 'Lisa', 'Jens'),
                         LName = c('Johnson', 'Johnson', 'Ericson', 'Peterson', 'Stephenson', 'Bond', 'Bond', 'Gustafsson'),
                         Age = c(34, 34, 40, 44, 44, 51, 51, 50),
                         Gender = c('M', 'M', 'F', 'M', 'M', 'F', 'F', 'M'))

# Remove duplicated rows from data frame
example_df.un <- example_df %>%
  distinct()

# Remove duplicate rows using the unique() function
examp_df <- unique(example_df)

# Drop duplicated columns
# In this case, there are no truly duplicated columns in the dataset,
# but if there were, we could drop them like this:
ex_df.un <- example_df[, !duplicated(as.list(example_df))]

# Drop duplicates with the distinct() function (from dplyr package)
ex_df.remove <- example_df %>%
  distinct()

# Remove duplicate rows based on the "Age" column, but keep all columns
ex_df_age <- example_df %>%
  distinct(Age, .keep_all = TRUE)

# Display the results
example_df.un  # Without any duplicated rows
examp_df       # Using unique()
ex_df.un       # Checking if any duplicated columns were dropped
ex_df.remove   # Dropping rows based on all columns
ex_df_age      # Removing duplicates based only on Age

```

End-of-File\
Pongsun B.\
2024-08-22
